Seismic parameter estimation and the Canadian crust

### 2013-09-21 paper/draftBP.tex
discontinuity discussion

\end{itemize}
%% -----------------------------------------------%%
\section{Data and methods}
Our analysis of bulk Canadian continental crust is based on estimates of crustal properties that can be accessed via seismic techniques, namely seismic wave velocities (or velocity ratio) and crustal thickness. This study draws on three sources for these crustal properties. The first and primary source is teleseismic receiver function (hereafter RF) analysis. The remaining two datasets are Crust 1.0, a widely published statistical average of similar regions with a global two degree resolution, and a compilation of pre-processed active source data (Mooney, 2012). Both Crust 1.0 and the active source compilation are used to qualify and compare with the primary dataset.

@@

%% -----------------------------------------------%%
\section{Results}

@@

\subsection{Crustal Discontinuities}

@@

We now consider the average crustal structure across Canada. We produce a depth profile for each station by stacking receiver functions along traveltime curves corresponding to a range of values in $H$ (0 to 50 km in increments of 100 metres) using the estimates for $R$ computed via the ZK + semblance method. The resulting profiles contain energy at depths corresponding to major velocity discontinuities. By assembling all depth profiles into a matrix, we may decompose the data via SVD into their principal components. Figure \ref{fig:conradPCA} contains the first five singular vectors or modes normalized to unit magnitude that account for 78\% of the variance in the profiles. Higher modes correspond to diminishing contributions to total variance. Each vector is normalized to unit magnitude. The first mode (blue) is effectively the average depth profile across all stations and captures 50\% of the variance in the dataset. It features a maximum near 38.7 km that may be taken as an average Moho depth. The second and third modes are dominated by a high amplitude shallow structures (e.g. sedimentary layers) within the top 10 km and Moho variability between 30 and 46 km . The fourth and fifth modes account for a combined 11\% of the variance in the dataset, contain energy at upper-mid crustal depths in the 10 km to 20 km range. All 5 modes display markedly lower energy within a depth window roughly between 20 km and 30 km.

@@

Seismic reflectivity of the continental crust is a valuable constraint on the processes governing crustal evolution. Research into the Mohoveric discontinuity and crustal thickness is ongoing and providing new insights into ...

1. introductory paragraph along the lines of the one you did for crustal composition, identifying the
importance of 1-D crustal structure for models of evolution. Introduce Conrad discontinuity and
general seismic reflection subdivision of crust into more transparent upper half and reflective
lower half, ie, summarize basic ideas from Brown EOS article on Conrad and Brocher and Mooney
GJRAS article

@@

>>>>

\end{itemize}
%% -----------------------------------------------%%

\section{Data and methods}
Our analysis of bulk Canadian continental crust is based on estimates of crustal properties that can be accessed via seismic techniques, namely seismic wave velocities (or velocity ratio) and crustal thickness. This study draws on three sources for these crustal properties. The first and primary source is teleseismic receiver function (hereafter RF) analysis. The remaining two datasets are Crust 1.0, a widely published statistical average of similar regions with a global two degree resolution, and a compilation of pre-processed active source data (Mooney, 2012). Both Crust 1.0 and the active source compilation are used to qualify and compare with the primary dataset.

@@

%% -----------------------------------------------%%

\section{Results}

@@

\subsection{Crustal Discontinuities} \label{section:discontinuities}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{discontinuities}
  \caption{}
  \label{fig:disc}
\end{figure}

We now consider the average crustal structure across Canada. We produce a depth profile for each station by stacking receiver functions along traveltime curves corresponding to a range of values in $H$ (0 to 50 km in increments of 100 metres) using the estimates for $R$ computed via the ZK + semblance method. The resulting profiles contain energy at depths corresponding to major velocity discontinuities and are shown in figure \ref{fig:disc} arranged by increasing crustal thickness. By assembling all depth profiles into a matrix, we may decompose the data via SVD into their principal components. Figure \ref{fig:conradPCA} contains the first five singular vectors or modes normalized to unit magnitude that account for 78\% of the variance in the profiles. Higher modes correspond to diminishing contributions to total variance. Each vector is normalized to unit magnitude. The first mode (blue) is effectively the average depth profile across all stations and captures 50\% of the variance in the dataset. It features a maximum near 38.7 km that may be taken as an average Moho depth. The second and third modes are dominated by a high amplitude shallow structures (e.g. sedimentary layers) within the top 10 km and Moho variability between 30 and 46 km . The fourth and fifth modes account for a combined 11\% of the variance in the dataset, contain energy at upper-mid crustal depths in the 10 km to 20 km range. All 5 modes display markedly lower energy within a depth window roughly between 20 km and 30 km.

@@

Seismic reflectivity of the continental crust is a valuable constraint on the processes governing crustal evolution. Early efforts to characterize continental crust revolved around a slab model with a lower velocity upper crust and a higher velocity lower crust [Conrad, 1925; Jeffreys, 1926]. Evidence for this model was originally put forth by Conrad [1925] who suggested observable P arrivals were propogating along a mid-crustal interface similar to the deeper crust-mantle interface imaged by Andrija Mohorovičić [Litak and Brown, 1989]. Both the crustal slab model and the Conrad Discontinuity, as the mid-crustal interface became known, persisted in the literature but were increasingly criticized as being too simplistic, in the case of the slab model, or, for the Conrad Discontinuity, lacking consistent evidence [Litak and Brown, 1989]. The development of reflection seismology allowed workers to further argue for a complex heterogenous crust and propose a model whereby smooth random lateral inhomogenities superimposed on a linearly increasing velocity gradient could produce seismic responses similar to that of a simple two layer model [Litak and Brown, 1989; Mereu and Ojo, 1981].

Despite the fading support for the Conrad Discontinuity as a discrete geological phenomena the generalization of the crust into seismically distinguishable upper and lower components continued. Mooney and Brocher [1987] argued for a model of the continental crust seperated into a seismically reflective lower crust overlain by transparent upper crust. Their physical interpretion is of a poorly laminated upper crust dominated by features of short length scale contrasted with a lower crust composed of thin (100 metre) lamellae of significant length scales. Supporting evidence for a reflective lower crust [Ito, 1998; Mayer, 1996; Meissner et. al., 2006] indicate that this is likely regional and representational of extented as well as thicker warmer crusts.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{discontinuityInterp}
  \caption{Left: Vertical station depth profiles interpolated to a common Moho. Right: Normalized stack of depth profiles, grey line shows mean}
  \label{fig:discInterp}
\end{figure}

The lower energy noted in Section \ref{section:discontinuities} in the 20 to 30 km depth window for the station profiles given in figure \ref{fig:disc} provides some evidence for seismically distinguishable crustal layers. To investigate this further we stretch and interpolate the profiles so that the Moho of each profile lines up on a common axis (Figure \ref{fig:discInterp}, left) and then take the magnitude of each profile and stack (Figure \ref{fig:discInterp}, right). The energy of the stack is most pronounced near the surface, associated with sedimentary layers, and at the Moho. There is a marked decrease in energy in the bottom third of the profile.

(Jackson 2002 - strength of lithosphere - mirrors decrease in energy)

@@

[[ADD GRANITE DIORITE REFERENCE]]
